---
title: "Brain Multiomic Analysis -- PCA Map"
author: "Qiuyi Wu"
date: "`r Sys.Date()`"
output:
  html_document:
    toc: true
    toc_depth: 2
    toc_float: true
    code_folding: hide
---

```{r setup, include=FALSE}
library(Hmisc)
library(sRDA)
library(PMA)
library(corrplot)
library(dplyr)
library(readxl)
library(ggplot2)
library(ggtext)
library(pheatmap)
library(gridExtra)
library(factoextra)
library(kableExtra)
library(reshape2)
library(pROC)
library(randomForest)
library(glmnet)
library(DT)
library(reshape2)
library(sva)
library(writexl)
library(ggrepel)
library(purrr)
library(tidyverse)
library(plyr)
library(DT)
library(htmltools)

knitr::opts_chunk$set(echo=TRUE, message=FALSE, cache=FALSE, warning=FALSE)
```

UPDATE on 08/21/2025:
Based on the CCA analysis spatial map we decide to use heldout sample 278 to replace original sample (AnalysisID) 10;  heldout sample 401 to replace original sample (AnalysisID) 229. Also remove sample  (AnalysisID) 96 and 188. 


# Bonnie A26

In this analysis, I proceed with 85% cutoff for batch correction (combat), the PCA analysis, mapping of PCA scores on the brain map. Totally 590 metabolites, and 388 metabolites have missing value less than 85%, and 202 metabolites have missingness larger than 85%. After discussion, we decide to remove those missing > 70% for TG metabolites, and remove those missing > 85% for other metabolites. The number of metabolites reduced from 590 to **328**. Combat is based on 403 samples (33 SPQC samples are removed after SPQC subtraction). PCA analysis is based on 269 samples (134 held out samples are removed). **Based on the result, we also decided to remove sample 244 (according to the 6/4/2025 group meeting)**. **Based on the downsteaming analysis, we also decided to remove sample 96 and 188 (according to the 8/20/2025 group meeting)**.

Steps for Combat:

1. Remove sample 244, 96, 188
1. For TG metabolites, remove those missing > 70%; for other metabolites, remove those missing > 85%. 
2. Impute LOD/2
3. Dilution factor
4. Remove 3 SPQC outliers
5. Log transform
6. Subtract SPQC
7. Combat
8. PCA analysis (remove HeldOut_AnalysisIDs in Mapped_Samples sheet)


```{r load metabolomic A26}
load("../../../data/data_A22.RData")
meta_met.a26 <- read_excel("../../../data/10412-Q500 Data.xlsx", sheet = "1. 10412 Raw Data", skip=1)
met_data.a26 <- meta_met.a26[!meta_met.a26$`Sample identification` %in% c(grep("Q500\\(XL\\)_QC", meta_met.a26$`Sample identification`, value = TRUE)), ] # Analysis ID
a26.met.sampletype <- ifelse(grepl("SPQC", met_data.a26$`Sample description`), "SPQC", "Non-SPQC")
met_data.a26 <- cbind(met_data.a26$`Sample description`, a26.met.sampletype, met_data.a26$`Submission name`, met_data.a26[, -c(1:11)])
colnames(met_data.a26)[1:3] <- c("Sample", "sampleType", "Plate")
# Remove all the ratios/derived features before any analysis, 630 metabolites left
met_data.a26 <- met_data.a26[, 1:633] 
features <- met_data.a26[1,-(1:3)]
met_data.a26 <- data.frame(met_data.a26[-c(1:14),])
colnames(met_data.a26)[-(1:3)] <- features

# Remove sample 244
met_data.a26 <- met_data.a26[!met_data.a26$Sample == "244",]
# Remove sample 96
met_data.a26 <- met_data.a26[!met_data.a26$Sample == "96",]
# Remove sample 188
met_data.a26 <- met_data.a26[!met_data.a26$Sample == "188",]

# Replace all occurrences of "<LOD" with NA in the entire dataset
met_data.a26[met_data.a26 == "<LOD"] <- NA
# Convert all elements in met_data.a26 from character to numeric
met_data.a26[,-c(1:3)] <- as.data.frame(lapply(met_data.a26[,-c(1:3)], function(x) as.numeric(as.character(x))))
# remove 100% missing metabolites
met_data.a26 <- met_data.a26[, colMeans(is.na(met_data.a26)) < 1] # 40 metabolites with 100% missingness are removed, we have 590 metabolites left. 
```



```{r imput}
# Identify which columns are TG metabolites
tg_metabolites <- grepl("^TG", colnames(met_data.a26))
# Apply different missingness cutoffs
impute_cols <- colMeans(is.na(met_data.a26)) <= ifelse(tg_metabolites, 0.70, 0.85)
# Split data based on cutoff criteria
met_data.a26.impute <- met_data.a26[, impute_cols]
met_data.a26.binary <- met_data.a26[, !impute_cols]


# impute metabolites 
met_data.a26.impute[,-c(1:3)] <- lapply(met_data.a26.impute[,-c(1:3)], function(col) {
  if (is.numeric(col)) {
    min_value <- min(col, na.rm = TRUE)
    col[is.na(col)] <- min_value / 2
  }
  return(col)
})
features <- colnames(met_data.a26.impute)[-c(1:3)]

# binary metabolites 
met_data.a26.binary <- ifelse(is.na(met_data.a26.binary), 0, 1)

# need to compare A22 and A26 scale -- dilution factor 
met_data.a26.unscaling <- met_data.a26.impute
met_data.a26.unscaling[,-c(1:3)] <- met_data.a26.impute[,-c(1:3)]*0.1

# removed 3 SPQC outliers
met_data.a26.unscaling <- met_data.a26.unscaling[-c(327:329), ]
met_data.a26.binary <- met_data.a26.binary[-c(327:329), ]

# log2 transformation
met_data.a26.unscaling[,-c(1:3)] <- log2(met_data.a26.unscaling[,-c(1:3)])
```


- For A26 plate 5, 3 SPQC outlier samples are removed.
- We average SPQC samples in each plate, and subtract the SPQC sample metabolites’ mean by plate. 
- Then we correct for batch effects in the SPQC adjsuted data using the `ComBat` function from the `sva` package.


```{r a26 spqc}
# substract SPQC average by plate
met_data.a26.spqcScaled <- array(dim = c(1,ncol(met_data.a26.unscaling)))
colnames(met_data.a26.spqcScaled) <- colnames(met_data.a26.unscaling)
met.plates <- unique(met_data.a26.unscaling$Plate)

for (plate in met.plates) {
  plate_data <- met_data.a26.unscaling[met_data.a26.unscaling$Plate == plate, ]
  spqc_means <- colMeans(plate_data[plate_data$sampleType == "SPQC", -c(1:3)])
  plate_data_spqcScaled <- plate_data
  plate_data_spqcScaled[, -c(1:3)] <- sweep(plate_data[, -c(1:3)], 2, spqc_means, FUN = "-")
  met_data.a26.spqcScaled <- rbind(met_data.a26.spqcScaled, plate_data_spqcScaled)
}
met_data.a26.spqcScaled <- met_data.a26.spqcScaled[-1,]

# remove SPQC samples, 403 samples
met_data.a26.spqcScaled <- met_data.a26.spqcScaled[met_data.a26.spqcScaled$sampleType != "SPQC", ]
rownames(met_data.a26.spqcScaled) <- met_data.a26.spqcScaled$Sample
```



```{r a26 combat}
# ComBat Batch correction
metabolite <- as.matrix(t(met_data.a26.spqcScaled[,-c(1:3)]))
met.batch <- as.factor(met_data.a26.spqcScaled$Plate)
# Apply ComBat to remove batch effect
combat_corrected <- ComBat(dat = metabolite, batch = met.batch, mod = NULL, par.prior = TRUE, prior.plots = FALSE)
df_corrected <- t(combat_corrected)

a26.met.data <- data.frame(cbind(met_data.a26.spqcScaled[,c(1:3)], df_corrected))
colnames(a26.met.data) = colnames(met_data.a26.spqcScaled)

# library(openxlsx)
# write.xlsx(a26.met.data,"../../data/PostCombat_A26_Metabolite_beforeHeldout.xlsx")

# PCA + PLOT
#df_corrected <- df_corrected[, which(var_values > 0)]  # Keep nonzero variance columns
pca_met_corrected <- prcomp(df_corrected, scale. = TRUE)
pca_data <- data.frame(pca_met_corrected$x)  # PCA scores
pca_data$plate <- met_data.a26.spqcScaled$Plate
pca_data$sample_type <- met_data.a26.spqcScaled$sampleType
pca_data$color_indicator <- ifelse(pca_data$sample_type == "SPQC", "black", pca_data$plate)
pca_data$sample <- as.numeric(met_data.a26.spqcScaled$Sample)

```

