---
title: "Brain Multiomic Analysis -- A26 Proteomics Data Batch Effect"
author: "Qiuyi Wu"
date: "`r Sys.Date()`"
output:
  html_document:
    toc: true
    toc_depth: 2
    toc_float: true
    code_folding: hide
---

```{r setup, include=FALSE}
library(Hmisc)
library(sRDA)
library(PMA)
library(corrplot)
library(dplyr)
library(readxl)
library(ggplot2)
library(ggtext)
library(pheatmap)
library(gridExtra)
library(factoextra)
library(kableExtra)
library(reshape2)
library(pROC)
library(randomForest)
library(glmnet)
library(DT)
library(reshape2)
library(sva)

knitr::opts_chunk$set(echo=TRUE, message=FALSE, cache=FALSE, warning=FALSE)

```

# Bonnie A26 
For the Bonnie data, to visualize batch effect, produce a PCA plot on the proteomics data, using samples from all plates including SPQC samples. Color the samples by plate and point shape by sample type (SPQC or not).



## Proteomics Data

```{r load proteomics A26}
load("../../data/Adjusted_Protein_A26.RData")
pro.comb.num <- as.data.frame(lapply(pro.a26.adj[,-c(1:3)], function(x) as.numeric(as.character(x))))

```

Extract the data from `Bonnie Proteomics with QC variables.txt`. There are totally `r dim(pro.a26.adj)[1]` samples, and `r dim(pro.a26.adj)[2]-3` proteins.


In the table below, I showed the number of missing values in each protein, and I use LOD/2 to impute these missing values before conducting PCA. After imputing missing data, I also scaled the data to match it with the unit in A22. 

The data processing order: log2 transformation -> impute missing values -> scale

### Proteomics PCA plot 

```{r combine proteomics A26 PCA}
# Check the missingness in the variables:
missingNumber.pro <- data.frame(Protein= colSums(is.na(pro.comb.num)))

datatable(missingNumber.pro,
          options = list(pageLength = 10),
          caption = htmltools::tags$caption(
            style = 'caption-side: top; text-align: left; font-size: 120%; font-weight: bold;',
            "Number of Missing values in each protein"))



# LOD/2 imputation for the missingness 
pro.comb.num[] <- apply(pro.comb.num, 2, function(x) {
  x[is.na(x)] <- min(x, na.rm = TRUE)/2
  return(x)
})


# Scale the data to make it unified with A22: A26 raw data * 0.1
# which means the log2 transformed processed data should be added term log2(0.1) 
# log2(A26 raw data * 0.1) = log2(0.1) + log2(raw data)

pro.comb.num <- pro.comb.num + log2(0.1)
pca_pro <- prcomp(pro.comb.num, scale. = TRUE)
pca_data <- data.frame(pca_pro$x[,1:2])  # PCA scores
pca_data$plate <- pro.a26.adj$pro.plate
pca_data$sample_type <- pro.a26.adj$pro.sampletype

outlier_ids <- which(abs(pca_data$PC1) > 300)
   

# Plot PCA using ggplot
ggplot(pca_data, aes(x=PC1, y=PC2, color=plate, shape=sample_type)) +
  geom_point(size=2) +
  labs(title="PCA Plot of Proteomics Data",
       x="PC 1",
       y="PC 2") +
  theme_minimal() +
  scale_color_manual(values=c("red", "blue", "green", "purple", "orange", "pink")) +
  scale_shape_manual(values=c(16, 17))  + # Shape 16 for SPQC, 17 for non-SPQC
    geom_text(data = pca_data[outlier_ids, ], 
              aes(x = pca_data[outlier_ids, 1], 
                  y = pca_data[outlier_ids, 2], 
                  label = pro.a26.adj$pro.sample[outlier_ids]),
              hjust = -0.4, vjust = 0.5, size = 4, color = "blue")




pca_data$color_indicator <- ifelse(pca_data$sample_type == "SPQC", "black", pca_data$plate)

ggplot(pca_data, aes(x=PC1, y=PC2, color=color_indicator, shape=sample_type)) +
  geom_point(size=2) +
  labs(title="PCA Plot of Proteomics Data (SPQC samples are colored black)",
       x="PC 1",
       y="PC 2") +
  theme_minimal() +
  scale_color_manual(values=c("black", "red", "blue", "green", "purple", "orange", "pink")) +  # Set black for SPQC
  scale_shape_manual(values=c(16, 17))




# 
# loadings_df <- as.data.frame(pca_result$rotation[, 1:2])  # PC1 and PC2
# loadings_df$PG.ProteinAccessions <- rownames(loadings_df)
# 
# # Step 6: Merge with precursor count and flag low-precursor proteins
# loadings_annot <- loadings_df %>%
#   left_join(precursor_info, by = "PG.ProteinAccessions") %>%
#   mutate(low_precursor = ifelse(PG.NrOfPrecursorsUsedForQuantification < 6, "Low (<6)", "Sufficient (≥6)"))
# 
# # Step 7: Plot
# ggplot(loadings_annot, aes(x = PC1, y = PC2, color = low_precursor)) +
#   geom_point(alpha = 0.7) +
#   theme_minimal() +
#   scale_color_manual(values = c("Low (<6)" = "red", "Sufficient (≥6)" = "black")) +
#   labs(
#     title = "PCA Loadings by Number of Precursors",
#     x = "PC1", y = "PC2",
#     color = "Precursor Count"
#   )


```


Now we remove "ID117530" sample in Plate 1, which has NO precursors used for quantification. And re-run the analysis.


```{r remove outlier}
# remove outlier "ID117530" sample
pro.a26.adj <- pro.a26.adj[pro.a26.adj$pro.sample != "ID117530",]
pro.comb.num <- as.data.frame(lapply(pro.a26.adj[,-c(1:3)], function(x) as.numeric(as.character(x))))


missingNumber.pro <- data.frame(Protein= colSums(is.na(pro.comb.num)))

datatable(missingNumber.pro,
          options = list(pageLength = 10),
          caption = htmltools::tags$caption(
            style = 'caption-side: top; text-align: left; font-size: 120%; font-weight: bold;',
            "Number of Missing values in each protein"))



# LOD/2 imputation for the missingness 
pro.comb.num[] <- apply(pro.comb.num, 2, function(x) {
  x[is.na(x)] <- min(x, na.rm = TRUE)/2
  return(x)
})


# Scale the data to make it unified with A22: A26 raw data * 0.1
# which means the log2 transformed processed data should be added term log2(0.1) 
# log2(A26 raw data * 0.1) = log2(0.1) + log2(raw data)

pro.comb.num <- pro.comb.num + log2(0.1)

pca_pro <- prcomp(pro.comb.num, scale. = TRUE)
pca_data <- data.frame(pca_pro$x[,1:2])  # PCA scores
pca_data$plate <- pro.a26.adj$pro.plate
pca_data$sample_type <- pro.a26.adj$pro.sampletype

outlier_ids <- which(abs(pca_data$PC1) >120)
   

# Plot PCA using ggplot
ggplot(pca_data, aes(x=PC1, y=PC2, color=plate, shape=sample_type)) +
  geom_point(size=2) +
  labs(title="PCA Plot of Proteomics Data",
       x="PC 1",
       y="PC 2") +
  theme_minimal() +
  scale_color_manual(values=c("red", "blue", "green", "purple", "orange", "pink")) +
  scale_shape_manual(values=c(16, 17))  + # Shape 16 for SPQC, 17 for non-SPQC
    geom_text(data = pca_data[outlier_ids, ], 
              aes(x = pca_data[outlier_ids, 1], 
                  y = pca_data[outlier_ids, 2], 
                  label = pro.a26.adj$pro.sample[outlier_ids]),
              hjust = -0.4, vjust = 0.5, size = 4, color = "blue")




pca_data$color_indicator <- ifelse(pca_data$sample_type == "SPQC", "black", pca_data$plate)

ggplot(pca_data, aes(x=PC1, y=PC2, color=color_indicator, shape=sample_type)) +
  geom_point(size=2) +
  labs(title="PCA Plot of Proteomics Data (SPQC samples are colored black)",
       x="PC 1",
       y="PC 2") +
  theme_minimal() +
  scale_color_manual(values=c("black", "red", "blue", "green", "purple", "orange", "pink")) +  # Set black for SPQC
  scale_shape_manual(values=c(16, 17))


```



### Proteomics PCA plot in Each Plate

For quality control purposes, we also perform PCA on each plate separately and produce a similar scatter plot.

I also show the outliers based on Mahalanobis distance and identify them in the corresponding PC plots (red dots). 

PC coordinates of each plate are shown in the table below.


```{r pro each plate}

pro.plates <- unique(pro.a26.adj$pro.plate)
tb <- vector("list", 6)

pro.comb.complete_spqcScaled <- array(dim = c(1,ncol(pro.a26.adj)-3))
colnames(pro.comb.complete_spqcScaled) <- colnames(pro.comb.num)
# Loop through each plate and perform PCA
for(plate in pro.plates) {
  # Subset data for the current plate
  plate_data <- pro.comb.num[pro.a26.adj$pro.plate == plate, ]
  identifier <- pro.a26.adj[pro.a26.adj$pro.plate == plate, 1:3]
  
  # subtract average SPQC by plate
  spqc_means <- colMeans(plate_data[identifier$pro.sampletype == "SPQC",], na.rm = T)
  plate_data_spqcScaled <- sweep(plate_data, 2, spqc_means, FUN = "-")
  pro.comb.complete_spqcScaled <- rbind(pro.comb.complete_spqcScaled, plate_data_spqcScaled)
  
  # Remove columns with zero variance (constant columns)
  plate_data_clean <- plate_data[, apply(plate_data, 2, var) > 0]
  pca_result <- prcomp(plate_data_clean, scale. = TRUE)
  
  # Center the SPQC samples to (0, 0) by subtracting the mean of the first principal component (PC1) and second principal component (PC2)
  scaled_pca_result <- scale(pca_result$x[, 1:2], center = TRUE, scale = FALSE)
  
  pca_data <- data.frame(scaled_pca_result)  # PCA scores
  pca_data$plate <- identifier$pro.plate
  pca_data$SPQC <- identifier$pro.sampletype
  pca_data$sampleID <- identifier$pro.sample

  
  # Calculate Mahalanobis distance for each sample in PCA space
  pca_scores <- scaled_pca_result
  centroid <- colMeans(pca_scores)  
  dist_mahalanobis <- apply(pca_scores, 1, function(x) dist(rbind(centroid, x)))
  
  # Identify outliers 
  outlier_threshold <- quantile(dist_mahalanobis, 0.9)
  outlier_ids <- which(dist_mahalanobis > outlier_threshold)


  # Plot PCA using ggplot
  p <- ggplot(pca_data, aes(x=PC1, y=PC2, color=SPQC)) +
    geom_point(size=2) +
    geom_point(data = pca_data[outlier_ids, ], aes(x = pca_data[outlier_ids, 1], y = pca_data[outlier_ids, 2]), color = "blue", size = 2) +  
    labs(title=paste("PCA Plot for Plate:", plate),
         x="PC 1",
         y="PC 2") +
    theme_minimal() +
    scale_color_manual(values=c("black", "red")) +
    geom_text(data = plate_data[outlier_ids, ],
              aes(x = pca_data[outlier_ids, 1],
                  y = pca_data[outlier_ids, 2],
                  label = pca_data$sampleID[outlier_ids]),
              hjust = 0.5, vjust = 2, size = 2, color = "blue")
  print(p)  
  
  
  
  # Create a data frame for PC coordinates -- original one before centralized 
  pc_coords <- data.frame(round(pca_result$x[, 1:2], 3))
  pc_coords$sample_id <- pca_data$sampleID 
  pc_coords$color_indicator <- pca_data$SPQC
  pc_coords$is_outlier <- ifelse(rownames(pc_coords) %in% rownames(plate_data[outlier_ids,]), TRUE, FALSE) 
  
  tb[[as.numeric(substr(plate, 6,7))]] <- pc_coords
}

datatable(tb[[1]],
          options = list(pageLength = 10),
          caption = htmltools::tags$caption(
            style = 'caption-side: top; text-align: left; font-size: 120%; font-weight: bold;',
            "PC coordinates for Plate 1"))

datatable(tb[[2]],
          options = list(pageLength = 10),
          caption = htmltools::tags$caption(
            style = 'caption-side: top; text-align: left; font-size: 120%; font-weight: bold;',
            "PC coordinates for Plate 2"))

datatable(tb[[3]],
          options = list(pageLength = 10),
          caption = htmltools::tags$caption(
            style = 'caption-side: top; text-align: left; font-size: 120%; font-weight: bold;',
            "PC coordinates for Plate 3"))

datatable(tb[[4]],
          options = list(pageLength = 10),
          caption = htmltools::tags$caption(
            style = 'caption-side: top; text-align: left; font-size: 120%; font-weight: bold;',
            "PC coordinates for Plate 4"))

datatable(tb[[5]],
          options = list(pageLength = 10),
          caption = htmltools::tags$caption(
            style = 'caption-side: top; text-align: left; font-size: 120%; font-weight: bold;',
            "PC coordinates for Plate 5"))

datatable(tb[[6]],
          options = list(pageLength = 10),
          caption = htmltools::tags$caption(
            style = 'caption-side: top; text-align: left; font-size: 120%; font-weight: bold;',
            "PC coordinates for Plate 6"))


```



```{r re-plot PCA protein}
pro.comb.complete_spqcScaled  <- pro.comb.complete_spqcScaled[-1,]


pca_pro <- prcomp(pro.comb.complete_spqcScaled, scale. = TRUE)
pca_data <- data.frame(pca_pro$x)  # PCA scores


pca_data$plate <- pro.a26.adj$pro.plate
pca_data$sample_type <- pro.a26.adj$pro.sampletype

pca_data$color_indicator <- ifelse(pca_data$sample_type == "SPQC", "black", pca_data$plate)

ggplot(pca_data, aes(x=PC1, y=PC2, color=color_indicator, shape=sample_type)) +
  geom_point(size=2) +
  labs(title="SPQC adjusted PCA Plot of Proteomics Data",
       x="PC 1",
       y="PC 2") +
  theme_minimal() +
  scale_color_manual(values=c("black", "red", "blue", "green", "purple", "orange", "pink")) +  # Set black for SPQC
  scale_shape_manual(values=c(16, 17))  
```

### Proteomics SVA Batch Correction
We also tried to correct for batch effects in the data using the `ComBat` function from the `sva` package.

```{r pro sva}
protein_data <- as.matrix(t(pro.comb.num))
pro.batch <- as.factor(pro.a26.adj$pro.plate)

# Apply ComBat to remove batch effect
combat_corrected <- ComBat(dat = protein_data, batch = pro.batch, mod = NULL, par.prior = TRUE, prior.plots = FALSE)


pca_pro_corrected <- prcomp(t(combat_corrected), scale. = TRUE)
pca_data <- data.frame(pca_pro_corrected$x)  # PCA scores
pca_data$plate <- pro.a26.adj$pro.plate
pca_data$sample_type <- pro.a26.adj$pro.sampletype 
pca_data$color_indicator <- ifelse(pca_data$sample_type == "SPQC", "black", pca_data$plate)


outlier_ids <- which(abs(pca_data$PC1) > 300)

ggplot(pca_data, aes(x=PC1, y=PC2, color=color_indicator, shape=sample_type)) +
  geom_point(size=2) +
  labs(title="PCA Plot of Proteomics Data using [ComBat]",
       x="PC 1",
       y="PC 2") +
  theme_minimal() +
  scale_color_manual(values=c("black", "red", "blue", "green", "purple", "orange", "pink")) +  
  scale_shape_manual(values=c(16, 17))  +
    geom_text(data = pca_data[outlier_ids, ], 
              aes(x = pca_data[outlier_ids, 1], 
                  y = pca_data[outlier_ids, 2], 
                  label = pro.a26.adj$pro.sample[outlier_ids]),
              hjust = -0.4, vjust = 0.5, size = 4, color = "blue")


combat_corrected_unscaled <- combat_corrected 

```



```{r pro spqc sva}
protein_data <- as.matrix(t(pro.comb.complete_spqcScaled))

# Apply ComBat to remove batch effect
combat_corrected <- ComBat(dat = protein_data, batch = pro.batch, mod = NULL, par.prior = TRUE, prior.plots = FALSE)


pca_pro_corrected <- prcomp(t(combat_corrected), scale. = TRUE)
pca_data <- data.frame(pca_pro_corrected$x)  # PCA scores
pca_data$plate <- pro.a26.adj$pro.plate
pca_data$sample_type <- pro.a26.adj$pro.sampletype 
pca_data$color_indicator <- ifelse(pca_data$sample_type == "SPQC", "black", pca_data$plate)


outlier_ids <- which(abs(pca_data$PC1) > 300)
  
ggplot(pca_data, aes(x=PC1, y=PC2, color=color_indicator, shape=sample_type)) +
  geom_point(size=2) +
  labs(title="SPQC adjusted PCA Plot of Proteomics Data using [ComBat]",
       x="PC 1",
       y="PC 2") +
  theme_minimal() +
  scale_color_manual(values=c("black", "red", "blue", "green", "purple", "orange", "pink")) +  
  scale_shape_manual(values=c(16, 17))  +
    geom_text(data = pca_data[outlier_ids, ], 
              aes(x = pca_data[outlier_ids, 1], 
                  y = pca_data[outlier_ids, 2], 
                  label = pro.a26.adj$pro.sample[outlier_ids]),
              hjust = -0.4, vjust = 0.5, size = 4, color = "blue")


combat_corrected_scaled <- combat_corrected




# save data

# combat_corrected_scaled <- cbind( pro.a26.adj[,1:3], t(combat_corrected_scaled))
# combat_corrected_unscaled <- cbind( pro.a26.adj[,1:3], t(combat_corrected_unscaled))
# library(openxlsx)
# write.xlsx(
#   list("combat_corrected_unscaled" = combat_corrected_unscaled, "combat_corrected_scaled" = combat_corrected_scaled),
#   file = "PostCombat_A26_Protein.xlsx"
# )
```

We decide NOT to subtract the SPQC (plate-wise) before batch correction.





