---
title: "Sample Size Study for A26"
author: "Qiuyi Wu"
date: "`r Sys.Date()`"
output:
  html_document:
    toc: true
    toc_depth: 2
    toc_float: true
    code_folding: hide
    
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, fig_caption = TRUE, warning = FALSE, message = FALSE)
```




```{r lib, warning=FALSE, message=FALSE}
rm(list=ls())

# for data formatting
library(tidyverse)
library(dplyr)
library(plyr)
library(data.table)
library(httr)
library(reshape2)
library(DT)

# for plotting
library(ggplot2)
library(ggpubr)
library(cowplot)
library(gridExtra)
library(pheatmap)
library(ggfortify)
library(readxl)
```




```{r preparation}
heldout_updated <- readxl::read_xlsx("data/Heldout_Updated_a26.xlsx") %>% as.data.frame() %>% .[, c(2:5)]

id_map <- heldout_updated %>%
  select(Mapped_AnalysisID, HeldOut_AnalysisID1, HeldOut_AnalysisID2, finalID) %>%
  tidyr::pivot_longer(
    cols = c(Mapped_AnalysisID, HeldOut_AnalysisID1, HeldOut_AnalysisID2),
    values_to = "ID"
  ) %>%
  filter(!is.na(ID)) %>%
  distinct(ID, .keep_all = TRUE) %>%
  select(ID, finalID)

lookup <- setNames(id_map$finalID, id_map$ID)



###########


# finally 267 samples
a26.met.data <- as.data.frame(read_excel("data/PostCombat_A26_Metabolite_beforeHeldout.xlsx"))
a26.pro.data <- as.data.frame(read_excel("data/PostCombat_A26_Protein.xlsx", sheet = "combat_corrected_unscaled"))
# convert Duke ID to Analysis ID
dukeid.df <- data.frame(read_excel("data/A26_Bonnie_Flatmap_Grid_20260315.xlsx", sheet = "Proteome_DukeIDs"))[, -1]
colnames(a26.pro.data)[1] <- "DukeID"
a26.pro.data$DukeID <- as.numeric(substr(a26.pro.data$DukeID, 3,8))
# Join and replace DukeID with AnalysisID
a26.pro.data <- a26.pro.data %>%
  left_join(dukeid.df, by = "DukeID") %>%
  mutate(DukeID = AnalysisID) %>%       
  select(-AnalysisID)                   

heldout_updated <- as.data.frame(read_excel("data/Heldout_Updated_a26.xlsx"))

a26.met.data <- a26.met.data[a26.met.data$Sample %in% unique(heldout_updated$finalID ), ]
a26.pro.data <- a26.pro.data[a26.pro.data$DukeID %in% unique(heldout_updated$finalID ), ]

rownames(a26.met.data) <- a26.met.data$Sample
rownames(a26.pro.data) <- a26.pro.data$DukeID




############
load("data/sPCA_a26metCluster.RData")

nb <- spca_delaunay_metCluster$lw$neighbours  # adjacency list
# convert to binary adjacency matrix
n <- length(nb)
nbr1.a26 <- matrix(0, n, n)
### 1st neighbor matrix ###
for (i in 1:n) {
  nbr1.a26[i, nb[[i]]] <- 1
}

rownames(nbr1.a26)=colnames(nbr1.a26) = rownames(spca_delaunay_metCluster$li)

### 2nd neighbor matrix ###
nbr2.a26 <- 1*(nbr1.a26%*%nbr1.a26!=0) - nbr1.a26
nbr2.a26[is.na(nbr1.a26)] <- 0


### 3rd neighbor matrix ###
nbr3.a26 <- 1*(nbr1.a26%*%nbr1.a26%*%nbr1.a26!=0) - nbr1.a26 - nbr2.a26
nbr3.a26[is.na(nbr3.a26)] <- 0


######
rename_ids <- function(ids, lookup) {
  ids_chr <- as.character(ids)
  mapped <- lookup[ids_chr]
  ifelse(!is.na(mapped), mapped, ids_chr)
}

new_ids <- rename_ids(rownames(nbr1.a26), lookup)

rownames(nbr1.a26) <- new_ids
colnames(nbr1.a26) <- new_ids
rownames(nbr2.a26) <- new_ids
colnames(nbr2.a26) <- new_ids
rownames(nbr3.a26) <- new_ids
colnames(nbr3.a26) <- new_ids


# in preparation (after building neighbors)
nbr1.full <- nbr1.a26
nbr2.full <- nbr2.a26
nbr3.full <- nbr3.a26


```

The correlation method used in this report is "spearman" method.

# Metabolomic data
```{r a26 met}
a26.met <- a26.met.data
## metabolite columns 
met.a26.cols <- setdiff(colnames(a26.met),c("Sample", "sampleType", "Plate"))

## matrix of metabolites
X <- as.matrix(a26.met[, met.a26.cols])
rownames(X) <- a26.met$Sample


###
ids_chr <- as.character(rownames(X))
mapped <- lookup[ids_chr]
new_ids <- ifelse(!is.na(mapped), mapped, ids_chr)

rownames(X) <- as.character(new_ids)
###


common_samples <- intersect(rownames(X), rownames(nbr1.a26))
X.sub.met <- X[common_samples, , drop = FALSE]

# reorder all neighbor matrices
nbr1.a26 <- nbr1.a26[common_samples, common_samples]
nbr2.a26 <- nbr2.a26[common_samples, common_samples]
nbr3.a26 <- nbr3.a26[common_samples, common_samples]

feature_mean <- colMeans(X.sub.met, na.rm = TRUE)
feature_sd   <- apply(X.sub.met, 2, sd, na.rm = TRUE)

summary(feature_mean)
summary(feature_sd)

hist(feature_sd, breaks = 50,
     main = "Distribution of metabolite SDs",
     xlab = "Standard deviation")

plot(feature_mean, feature_sd,
     xlab = "Mean",
     ylab = "SD",
     main = "Mean-SD relationship")
```

Here I used Z-score to make each metabolite column becomes mean 0, SD 1 across samples.

```{r scaling met}
X.sub.met <- scale(X.sub.met, center = TRUE, scale = TRUE)

# Pearson as default

avg_neighbor_cor_se <- function(X, nbr_mat) {
  idx <- which(nbr_mat == 1, arr.ind = TRUE)
  idx <- idx[idx[,1] < idx[,2], , drop = FALSE]

  cors <- apply(idx, 1, function(k) {
    cor(X[k[1], ], X[k[2], ], use = "pairwise.complete.obs")
  })

  c(mean = mean(cors, na.rm = TRUE),
    se   = sd(cors, na.rm = TRUE) / sqrt(sum(!is.na(cors))))
}

avg_neighbor_cor_se.spearman <- function(X, nbr_mat) {
  idx <- which(nbr_mat == 1, arr.ind = TRUE)
  idx <- idx[idx[,1] < idx[,2], , drop = FALSE]

  cors <- apply(idx, 1, function(k) {
    cor(X[k[1], ], X[k[2], ], use = "pairwise.complete.obs", method = "spearman")
  })

  c(mean = mean(cors, na.rm = TRUE),
    se   = sd(cors, na.rm = TRUE) / sqrt(sum(!is.na(cors))))
}


avg_neighbor_cor_raw.spearman <- function(X, nbr_mat) {
  idx <- which(nbr_mat == 1, arr.ind = TRUE)
  idx <- idx[idx[,1] < idx[,2], , drop = FALSE]

  cors <- apply(idx, 1, function(k) {
    cor(X[k[1], ], X[k[2], ],
        use = "pairwise.complete.obs", method = "spearman")
  })
  cors[!is.na(cors)]
}


cors.1 <- avg_neighbor_cor_raw.spearman(X.sub.met, nbr1.a26)
cors.2 <- avg_neighbor_cor_raw.spearman(X.sub.met, nbr2.a26)
cors.3 <- avg_neighbor_cor_raw.spearman(X.sub.met, nbr3.a26)
box_data <- list(cors.1, cors.2, cors.3)

df.a26.met <- data.frame(
  order = 1:3,
  mean = sapply(box_data, mean),
  se   = sapply(box_data, sd) / sqrt(sapply(box_data, length))
)

yl.met <- range(box_data)

plot(df.a26.met$order, df.a26.met$mean,
     xlim = c(0.5, 3.5),
     ylim = yl.met,
     type = "b", pch = 16,
     xaxt = "n",
     xlab = "Neighbor order",
     ylab = "Average metabolomic correlation", 
     main = "Decay of A26 Metabolomic similarity with neighbor order")

axis(1, at = 1:3, labels = c("1st", "2nd", "3rd"))

boxplot(box_data,
        add = TRUE,
        at = 1:3,
        boxwex = 0.45,
        outline = FALSE,
        col = rgb(0, 0, 0, 0.15),
        border = "gray40",
        axes = FALSE)


```




```{r map met}
map.df <- as.matrix(readxl::read_xlsx("data/A26_Bonnie_Flatmap_Grid_20260315.xlsx", sheet = "AnalysisID_Matrix", col_names = F))
map_chr <- as.character(map.df)
replacement <- lookup[map_chr]
map_chr <- ifelse(!is.na(replacement), replacement, map_chr)
map.df <- matrix(as.numeric(map_chr), nrow=nrow(map.df))



avg_neighbor_cor <- function(X, nbr_mat, method = "pearson", use = "pairwise.complete.obs") {
  n <- nrow(X)
  out <- rep(NA_real_, n)
  names(out) <- rownames(X)
  for (i in seq_len(n)) {
    nb_idx <- which(nbr_mat[i, ] != 0)  # neighbors of i
    if (length(nb_idx) == 0) {
      out[i] <- NA_real_
      next
    }
    cors <- vapply(
      nb_idx,
      function(j) suppressWarnings(cor(X[i, ], X[j, ], method = method, use = use)),
      numeric(1)
    )
    out[i] <- mean(cors, na.rm = TRUE)
  }
  out
}

# X.sub.met: samples x metabolites matrix with rownames = sample IDs
# nbr1.a26, nbr2.a26, nbr3.a26: samples x samples adjacency matrices

cor1.met <- avg_neighbor_cor(X.sub.met, nbr1.a26, method = "spearman")
cor2.met <- avg_neighbor_cor(X.sub.met, nbr2.a26, method = "spearman")
cor3.met <- avg_neighbor_cor(X.sub.met, nbr3.a26, method = "spearman")

cor.df <- data.frame(
  Sample = names(cor1.met),
  cor_nbr1 = as.numeric(cor1.met),
  cor_nbr2 = as.numeric(cor2.met),
  cor_nbr3 = as.numeric(cor3.met),
  stringsAsFactors = FALSE
)









paint_to_map <- function(id_map, values_named) {
  id_chr <- matrix(as.character(id_map), nrow=nrow(id_map), ncol=ncol(id_map))
  id_chr[id_chr %in% c("NA","NaN","nan","", "0")] <- NA_character_  # keep "0" only if it's a real sample

  out <- matrix(NA_real_, nrow=nrow(id_chr), ncol=ncol(id_chr))
  idx <- which(!is.na(id_chr), arr.ind = TRUE)
  m <- match(id_chr[idx], names(values_named))
  out[idx] <- values_named[m]
  out
}


cor1_map <- paint_to_map(map.df, setNames(cor1.met, as.character(names(cor1.met))))
cor2_map <- paint_to_map(map.df, setNames(cor2.met, as.character(names(cor2.met))))
cor3_map <- paint_to_map(map.df, setNames(cor3.met, as.character(names(cor3.met))))


plot_map <- function(mat, title = "") {

  df <- expand.grid(
    row = seq_len(nrow(mat)),
    col = seq_len(ncol(mat))
  )

  df$value <- as.vector(mat)

  ggplot(df, aes(x = col, y = row, fill = value)) +
    geom_raster() +
    scale_y_reverse() +
    scale_x_reverse() +
    coord_equal() +
    scale_fill_gradient2(low = "blue", mid = "white",
    high = "red",
    midpoint = 0,
    limits = c(-1, 1),
    na.value = "grey90"
  ) +
    labs(title = title, x = NULL, y = NULL, fill = "avg cor") +
    theme_minimal()
}

(p1 <- plot_map(cor1_map, "A26 Metabolites 1st Neighbors Correlation Map"))
(p2 <- plot_map(cor2_map, "A26 Metabolites 2nd Neighbors Correlation Map"))
(p3 <- plot_map(cor3_map, "A26 Metabolites 3rd Neighbors Correlation Map"))

```



```{r plot1}
library(patchwork)

p1 <- plot_map(cor1_map, "") +
  ggtitle("1st Neighbor") +
  theme(plot.title = element_text(hjust = 0.5))

p2 <- plot_map(cor2_map, "") +
  ggtitle("2nd Neighbor") +
  theme(plot.title = element_text(hjust = 0.5))

p3 <- plot_map(cor3_map, "") +
  ggtitle("3rd Neighbor") +
  theme(plot.title = element_text(hjust = 0.5))

(p1 | p2 | p3) +
  plot_annotation(
    title = "A26 Metabolites Correlation Map",
    theme = theme(plot.title = element_text(size = 26, hjust = 0.5))
  )
```




# Proteomic data

```{r a26 proteomic}
a26.pro <- a26.pro.data
colnames(a26.pro)[1] <- "pro.sample"

## proteomic columns 
pro.a26.cols <- setdiff(colnames(a26.pro),c("pro.sample", "pro.sampletype", "pro.plate"))

## matrix of proteins
X <- as.matrix(a26.pro[, pro.a26.cols])
rownames(X) <- a26.pro$pro.sample

###
ids_chr <- as.character(rownames(X))
mapped <- lookup[ids_chr]
new_ids <- ifelse(!is.na(mapped), mapped, ids_chr)

rownames(X) <- as.character(new_ids)
###

```


```{r a26 pro}
# in preparation (after building neighbors)
nbr1.a26 <- nbr1.full 
nbr2.a26 <- nbr2.full
nbr3.a26 <- nbr3.full 

common_samples <- intersect(rownames(X), rownames(nbr1.a26))
X.sub.pro <- X[common_samples, , drop = FALSE]

# reorder all neighbor matrices
nbr1.a26 <- nbr1.a26[common_samples, common_samples]
nbr2.a26 <- nbr2.a26[common_samples, common_samples]
nbr3.a26 <- nbr3.a26[common_samples, common_samples]


feature_mean <- colMeans(X.sub.pro, na.rm = TRUE)
feature_sd   <- apply(X.sub.pro, 2, sd, na.rm = TRUE)

summary(feature_mean)
summary(feature_sd)

hist(feature_sd, breaks = 50,
     main = "Distribution of protein SDs",
     xlab = "Standard deviation")

plot(feature_mean, feature_sd,
     xlab = "Mean",
     ylab = "SD",
     main = "Mean-SD relationship")
```


We can see the scaling is needed for proteomic data to make the features comparable. I used Z-score to make each protein column becomes mean 0, SD 1 across samples.


```{r scaling}
X.z.pro <- scale(X.sub.pro, center = TRUE, scale = TRUE)

cors.1 <- avg_neighbor_cor_raw.spearman(X.z.pro, nbr1.a26)
cors.2 <- avg_neighbor_cor_raw.spearman(X.z.pro, nbr2.a26)
cors.3 <- avg_neighbor_cor_raw.spearman(X.z.pro, nbr3.a26)
box_data <- list(cors.1, cors.2, cors.3)

df.a26.pro <- data.frame(
  order = 1:3,
  mean = sapply(box_data, mean),
  se   = sapply(box_data, sd) / sqrt(sapply(box_data, length))
)

yl.pro <- range(box_data)

plot(df.a26.pro$order, df.a26.pro$mean,
     xlim = c(0.5, 3.5),
     ylim = yl.pro,
     type = "b", pch = 16,
     xaxt = "n",
     xlab = "Neighbor order",
     ylab = "Average Proteomic correlation", 
     main = "Decay of A26 Proteomic similarity with neighbor order")

axis(1, at = 1:3, labels = c("1st", "2nd", "3rd"))

boxplot(box_data,
        add = TRUE,
        at = 1:3,
        boxwex = 0.45,
        outline = FALSE,
        col = rgb(0, 0, 0, 0.15),
        border = "gray40",
        axes = FALSE)

```

```{r map pro}
cor1.pro <- avg_neighbor_cor(X.z.pro, nbr1.a26, method = "spearman")
cor2.pro <- avg_neighbor_cor(X.z.pro, nbr2.a26, method = "spearman")
cor3.pro <- avg_neighbor_cor(X.z.pro, nbr3.a26, method = "spearman")

cor.df <- data.frame(
  Sample = names(cor1.pro),
  cor_nbr1 = as.numeric(cor1.pro),
  cor_nbr2 = as.numeric(cor2.pro),
  cor_nbr3 = as.numeric(cor3.pro),
  stringsAsFactors = FALSE
)


cor1_map <- paint_to_map(map.df, setNames(cor1.pro, as.character(names(cor1.pro))))
cor2_map <- paint_to_map(map.df, setNames(cor2.pro, as.character(names(cor2.pro))))
cor3_map <- paint_to_map(map.df, setNames(cor3.pro, as.character(names(cor3.pro))))

(p1 <- plot_map(cor1_map, "A26 Proteins 1st Neighbors Correlation Map"))
(p2 <- plot_map(cor2_map, "A26 Proteins 2nd Neighbors Correlation Map"))
(p3 <- plot_map(cor3_map, "A26 Proteins 3rd Neighbors Correlation Map"))
```


# Protein Clusters

Use loading to aggregate proteins to clusters.

```{r pro cluster}
a26.pro.cluster <- readxl::read_xlsx("data/pro_cluster_contribution_a26.xlsx", sheet = 1) %>% data.frame()

make_sample_by_cluster_weighted_sum <- function(X.sub.pro, a26.pro.cluster) {
  X <- as.matrix(X.sub.pro)

  map <- a26.pro.cluster %>%
    transmute(
      Feature = as.character(Feature),
      Cluster = as.character(Cluster),
      Loading = as.numeric(Loading)
    ) %>%
    filter(Feature %in% colnames(X)) %>%
    filter(!is.na(Loading))

  feats_by_cluster <- split(map, map$Cluster)

  cluster_mat <- sapply(feats_by_cluster, function(df) {
    feats <- df$Feature
    w <- df$Loading
    # X[, feats] %*% w  => weighted sum per sample
    as.numeric(X[, feats, drop = FALSE] %*% w)
  })

  cluster_mat <- as.data.frame(cluster_mat, check.names = FALSE)
  cluster_mat
}

X.cluster.pro <- make_sample_by_cluster_weighted_sum(X.z.pro, a26.pro.cluster)
## matrix of protein clusters
X.cluster.pro <- as.matrix(X.cluster.pro)
rownames(X.cluster.pro) <- rownames(X.sub.pro)



cors.1 <- avg_neighbor_cor_raw.spearman(X.cluster.pro, nbr1.a26)
cors.2 <- avg_neighbor_cor_raw.spearman(X.cluster.pro, nbr2.a26)
cors.3 <- avg_neighbor_cor_raw.spearman(X.cluster.pro, nbr3.a26)
box_data <- list(cors.1, cors.2, cors.3)

df.a26.pro <- data.frame(
  order = 1:3,
  mean  = sapply(box_data, mean, na.rm = TRUE),
  se    = sapply(box_data, sd, na.rm = TRUE) / sqrt(sapply(box_data, function(x) sum(!is.na(x))))
)


yl.pro <- range(unlist(box_data), na.rm = TRUE)

plot(df.a26.pro$order, df.a26.pro$mean,
     xlim = c(0.5, 3.5),
     ylim = yl.pro,
     type = "b", pch = 16,
     xaxt = "n",
     xlab = "Neighbor order",
     ylab = "Average Cluster-level proteomic correlation",
     main = "Decay of A26 Proteomic Cluster similarity with neighbor order")

axis(1, at = 1:3, labels = c("1st", "2nd", "3rd"))

boxplot(box_data,
        add = TRUE,
        at = 1:3,
        boxwex = 0.45,
        outline = FALSE,
        col = rgb(0, 0, 0, 0.15),
        border = "gray40",
        axes = FALSE)

```


```{r map pro cluster}
cor1.pro <- avg_neighbor_cor(X.cluster.pro, nbr1.a26, method = "spearman")
cor2.pro <- avg_neighbor_cor(X.cluster.pro, nbr2.a26, method = "spearman")
cor3.pro <- avg_neighbor_cor(X.cluster.pro, nbr3.a26, method = "spearman")

cor.df <- data.frame(
  Sample = names(cor1.pro),
  cor_nbr1 = as.numeric(cor1.pro),
  cor_nbr2 = as.numeric(cor2.pro),
  cor_nbr3 = as.numeric(cor3.pro),
  stringsAsFactors = FALSE
)


cor1_map <- paint_to_map(map.df, setNames(cor1.pro, as.character(names(cor1.pro))))
cor2_map <- paint_to_map(map.df, setNames(cor2.pro, as.character(names(cor2.pro))))
cor3_map <- paint_to_map(map.df, setNames(cor3.pro, as.character(names(cor3.pro))))

(p1 <- plot_map(cor1_map, "A26 Protein Clusters 1st Neighbors Correlation Map"))
(p2 <- plot_map(cor2_map, "A26 Protein Clusters 2nd Neighbors Correlation Map"))
(p3 <- plot_map(cor3_map, "A26 Protein Clusters 3rd Neighbors Correlation Map"))
```


```{r plot2}
library(patchwork)

p1 <- plot_map(cor1_map, "") +
  ggtitle("1st Neighbor") +
  theme(plot.title = element_text(hjust = 0.5))

p2 <- plot_map(cor2_map, "") +
  ggtitle("2nd Neighbor") +
  theme(plot.title = element_text(hjust = 0.5))

p3 <- plot_map(cor3_map, "") +
  ggtitle("3rd Neighbor") +
  theme(plot.title = element_text(hjust = 0.5))

(p1 | p2 | p3) +
  plot_annotation(
    title = "A26 Protein Clusters Correlation Map",
    theme = theme(plot.title = element_text(size = 26, hjust = 0.5))
  )
```