---
title: "Cross-animal projection of A27-derived spCCA components onto A26"
author: "Qiuyi Wu"
date: "`r Sys.Date()`"
output:
  html_document:
    toc: true
    toc_depth: 2
    toc_float: true
    code_folding: hide
---

```{r setup, include=FALSE}
library(Hmisc)
library(sRDA)
library(PMA)
library(corrplot)
library(dplyr)
library(readxl)
library(ggplot2)
library(ggtext)
library(pheatmap)
library(gridExtra)
library(factoextra)
library(kableExtra)
library(reshape2)
library(deldir)
library(igraph)
library(spdep)
library(patchwork)
library(sRDA)
library(ggrepel)

knitr::opts_chunk$set(echo=TRUE, message=FALSE, cache=FALSE, warning=FALSE)
```



```{r omics map}
heldout_updated <- as.data.frame(read_excel("../data/Heldout_Updated_a26.xlsx"))
heldout_updated <- heldout_updated[, c(2:5)]
map.df <- data.frame(read_excel("../data/A26_Bonnie_Flatmap_Grid_20260315.xlsx", sheet = "AnalysisID_Matrix", col_names = F))
map.df <- as.matrix(sapply(map.df, as.numeric))
load("../data/a26.RData")


# Modify MAP based on HELDOUT: Build a lookup table of all IDs → finalID
id_map <- heldout_updated %>%
  select(Mapped_AnalysisID, HeldOut_AnalysisID1, HeldOut_AnalysisID2, finalID) %>%
  tidyr::pivot_longer(
    cols = c(Mapped_AnalysisID, HeldOut_AnalysisID1, HeldOut_AnalysisID2),
    values_to = "ID"
  ) %>%
  filter(!is.na(ID)) %>%
  distinct(ID, .keep_all = TRUE) %>%
  select(ID, finalID)
# Convert to named vector for fast replacement
lookup <- setNames(id_map$finalID, id_map$ID)

map_char <- as.character(map.df)               # all entries as character for indexing
replacement <- lookup[map_char]                 # yields NA when not found
final_vec <- ifelse(!is.na(replacement), replacement, map_char)  # keep original when not found
map.df<- matrix(as.numeric(final_vec), nrow = nrow(map.df), ncol = ncol(map.df))


load("../data/A26_PChclust.RData")
rownames(a26.pro.cluster$score) <- rownames(a26.met.cluster$score)

get_sample_coordinates <- function(block_matrix) {
  coords <- which(!is.na(block_matrix), arr.ind = TRUE)
  samples <- block_matrix[coords]
  xy_df <- data.frame(
    sample = as.character(samples),
    x = coords[, "col"],  # column index as x
    y = coords[, "row"]   # row index as y
  )
  return(xy_df)
}

map.df <- apply(map.df, 2, as.numeric)
# set 0 value as NA
map.df[map.df  == 0] <- NA

xy_coords <- get_sample_coordinates(apply(map.df, 2, rev))
xy_coords$sample <- as.character(xy_coords$sample)
# Reorder coordinates to match met_design
xy_ordered <- xy_coords[match(rownames(a26.met.cluster$score), xy_coords$sample), c("x", "y")]

# Subset both datasets to common samples
common_samples <- intersect(rownames(a26.met.cluster$score), unique(xy_coords$sample))
met_score_common <- a26.met.cluster$score[common_samples, , drop = FALSE]
pro_score_common <- a26.pro.cluster$score[common_samples, , drop = FALSE]

xy_coords_common <- xy_coords[match(common_samples, xy_coords$sample), c("x", "y")]
rownames(xy_coords_common) <- common_samples

#rownames(xy_ordered) <- xy_coords[match(rownames(a26.met.cluster$score), xy_coords$sample), "sample"]
# remove NA rows (where there's no match between two datasets)
#xy_ordered <- na.omit(xy_ordered)

xy_ordered <- xy_coords_common
```



```{r PCA}
#Function to detect outliers based on median ± k * MAD （detect outliers based on the median ± k × MAD (Median Absolute Deviation)）
mad_outliers <- function(x, k = 10) {
  med <- median(x, na.rm = TRUE)
  m <- mad(x, constant = 1, na.rm = TRUE)   # default constant=1
  out <- x[abs(x - med) > k * m]
  return(out)
}
# Metabolomics
pca_met_corrected <- prcomp(a26.met.data[,-c(1:3)], scale. = TRUE)
pca_met <- data.frame(pca_met_corrected$x)  # PCA scores
pca_met$plate <- a26.met.data$Plate
pca_met$sample_type <- a26.met.data$sampleType
pca_met$color_indicator <- ifelse(pca_met$sample_type == "SPQC", "black", pca_met$plate)
pca_met$sample <- as.numeric(a26.met.data$Sample)

# Detect outliers for PC1 and PC2
pc1_outliers <- mad_outliers(pca_met$PC1, k = 7)
pc2_outliers <- mad_outliers(pca_met$PC2, k = 7)
# Subset outliers
outlier_met <- pca_met[pca_met$PC1 %in% pc1_outliers | pca_met$PC2 %in% pc2_outliers, ]


ggplot(pca_met, aes(x=PC1, y=PC2, color=color_indicator)) +
  geom_point(size=2) +
  geom_text_repel(data=outlier_met, aes(label=sample), size=3, color="black") +  # Add outlier labels
  labs(title="PCA Plot of A26 Metabolomics Data After Heldout",
       x="PC 1",
       y="PC 2") +
  theme_minimal() +
  scale_color_manual(values=c("black", "red", "blue", "green", "purple", "orange", "pink")) +  # Set black for SPQC
  scale_shape_manual(values=c(16, 17))  



# Proteomics
pca_pro_corrected <- prcomp(a26.pro.data[,-c(1:3)], scale. = TRUE)
pca_pro <- data.frame(pca_pro_corrected$x)  # PCA scores
pca_pro$plate <- a26.pro.data$pro.plate 
pca_pro$sample_type <- a26.pro.data$pro.sampletype 
pca_pro$color_indicator <- ifelse(pca_pro$sample_type == "SPQC", "black", pca_pro$plate)
pca_pro$sample <- as.numeric(a26.pro.data$DukeID)

# Detect outliers for PC1 and PC2
pc1_outliers <- mad_outliers(pca_pro$PC1, k = 7)
pc2_outliers <- mad_outliers(pca_pro$PC2, k = 7)
# Subset outliers
outlier_pro <- pca_pro[pca_pro$PC1 %in% pc1_outliers | pca_pro$PC2 %in% pc2_outliers, ]


ggplot(pca_pro, aes(x=PC1, y=PC2, color=color_indicator)) +
  geom_point(size=2) +
  geom_text_repel(data=outlier_pro, aes(label=sample), size=3, color="black") +  # Add outlier labels
  labs(title="PCA Plot of A26 Proteomics Data After Heldout",
       x="PC 1",
       y="PC 2") +
  theme_minimal() +
  scale_color_manual(values=c("black", "red", "blue", "green", "purple", "orange", "pink")) +  # Set black for SPQC
  scale_shape_manual(values=c(16, 17))  


```




```{r spCCA cluster 1st component}
a26.met.cluster$score <- a26.met.cluster$score[rownames(a26.met.cluster$score) %in% rownames(xy_ordered), ]
a26.pro.cluster$score <- a26.pro.cluster$score[rownames(a26.pro.cluster$score) %in% rownames(xy_ordered), ]


# spca_delaunay_metCluster <- adegenet::spca(
#   x = a26.met.cluster$score,
#   xy = xy_ordered,
#   center = TRUE,
#   scale = TRUE,
#   scannf = FALSE,
#   nfposi = 3,
#   nfnega = 0
# )
# save(spca_delaunay_metCluster,  file = "sPCA_a26metCluster.RData")
load("../data/sPCA_a26metCluster.RData")

# Standardize both datasets (center + scale)
X <- a26.pro.cluster$score
Y <- a26.met.cluster$score

rownames(X) <- rownames(xy_ordered)

# Get coordinates and rownames
coords <- as.matrix(xy_ordered)
sample_names <- rownames(xy_ordered)



# lw is your listw object
nb <- spca_delaunay_metCluster$lw$neighbours  # adjacency list
# convert to binary adjacency matrix
n <- length(nb)
W_raw <- matrix(0, n, n)
for (i in 1:n) {
  W_raw[i, nb[[i]]] <- 1
}
# degree vector
d_vec <- rowSums(W_raw)
# diagonal degree matrix
D <- diag(d_vec)
# Unnormalized Laplacian
L <- D - W_raw

# Symmetric normalized Laplacian
L_sym <- diag(1 / sqrt(d_vec)) %*% L %*% diag(1 / sqrt(d_vec))
```


## Spatial CCA {.tabset}


To evaluate the reproducibility of the spatial canonical patterns identified in A27, we projected the A27-derived spCCA components onto the A26 data. Because spCCA was performed on cluster-level representations rather than the original feature matrices, we first mapped the A27 canonical loading vectors back to the original feature space. Specifically, for the proteomic data, the A27 spCCA loading matrix was multiplied by the A27 proteomic cluster loading matrix to obtain feature-level weights; the same procedure was applied to the metabolomic data. This yielded modality-specific feature weight matrices in the original proteomic and metabolomic spaces.

Next, we identified the set of features shared between A26 and A27 within each modality and restricted the projection to these common features. Using the A26 proteomic and metabolomic data matrices, we then computed projected scores by multiplying the A26 feature matrices by the A27-derived feature weights. In this way, each A26 sample received proteomic and metabolomic scores corresponding to the canonical components learned from A27.

Finally, for each component, these transferred scores were mapped back onto the cortical sample layout to generate spatial score maps for A26. This procedure allowed us to assess whether the cross-modality spatial patterns estimated in A27 were recapitulated in the independent A26 dataset. There are 9250 common proteins and 289 common metabolites. 


```{r sparse spCCA 2, message=FALSE, warning=FALSE, results='hide'}

# Inverse Laplacian (Moore–Penrose with small reg)
eig <- eigen(L_sym, symmetric = TRUE)
epsilon <- median(eig$values)  #max(eig$values) 
eta <- 1/epsilon
eig_inv_vals <- 1 / (eta*eig$values + 1)            # ifelse(eig$values > eps, 1 / eig$values, 0)
W_tilde <- eig$vectors %*% diag(eig_inv_vals) %*% t(eig$vectors)   # = L_sym^{-1}


##########

eig <- eigen(W_tilde)
W_half <- eig$vectors %*% diag(sqrt(pmax(eig$values, 0.01))) %*% t(eig$vectors)
W_half_inv  <- eig$vectors %*% diag(1/sqrt(pmax(eig$values, 0.01)), nrow=length(eig$values)) %*% t(eig$vectors)


X_tilde <- W_half %*% X
Y_tilde <- W_half %*% Y

Xtc <- X_tilde
Ytc <- Y_tilde



# load A27 score weights
load("../data/a27_score.RData")
load("../data/A27_PChclust.RData")
scca_res_a27 <- scca_res
ai27 <- scca_res_a27$u
bi27 <- scca_res_a27$v
pro_feat_wt_a27 <- a27.pro.cluster$loading %*% ai27
met_feat_wt_a27 <- a27.met.cluster$loading %*% bi27
rownames(pro_feat_wt_a27) = rownames(a27.pro.cluster$loading)
rownames(met_feat_wt_a27) = rownames(a27.met.cluster$loading)

common_pro <- intersect(rownames(a26.pro.cluster$loading), rownames(a27.pro.cluster$loading))
X26_pro_common <- as.matrix(a26.pro.data[, common_pro])
w27_pro_common <- pro_feat_wt_a27[common_pro,]
u_A26_from_A27 <- X26_pro_common %*% w27_pro_common
rownames(u_A26_from_A27) <- rownames(X26_pro_common)

common_met <- intersect(rownames(a26.met.cluster$loading), rownames(a27.met.cluster$loading))
X26_met_common <- as.matrix(a26.met.data[, common_met])
w27_met_common <- met_feat_wt_a27[common_met,]
v_A26_from_A27 <- X26_met_common %*% w27_met_common
rownames(v_A26_from_A27) <- rownames(X26_met_common)




for (i in 1:10) {
  u_scores <- u_A26_from_A27[,i]
  u_matrix <- matrix(u_scores[match(map.df, names(u_scores))], nrow = nrow(map.df))
  u_df <- reshape2::melt(apply(u_matrix, 2, rev), varnames = c("Row", "Column"), 
                         value.name = paste0("spCCA_U", i))

  p_u <- ggplot(u_df, aes_string("Column", "Row", fill = paste0("spCCA_U", i))) +
  geom_raster() + scale_x_reverse() + coord_equal() +
  scale_fill_viridis_c(option = "plasma", na.value = "grey90") +
  labs(title = paste0("Proteomic Score (U", i, ")"), x = NULL, y = NULL, fill = "Score") +
  theme_minimal() + theme( panel.grid = element_blank())


  v_scores <- v_A26_from_A27[,i]
  v_matrix <- matrix(v_scores[match(map.df, names(v_scores))], nrow = nrow(map.df))
  v_df <- reshape2::melt(apply(v_matrix, 2, rev), varnames = c("Row", "Column"), 
                         value.name = paste0("spCCA_V", i))
  
  p_v <- ggplot(v_df, aes_string("Column", "Row", fill = paste0("spCCA_V", i))) +
  geom_raster() + scale_x_reverse() + coord_equal() +
  scale_fill_viridis_c(option = "plasma", na.value = "grey90") +
  labs(title = paste0("Metabolomic Score (V", i, ")"), x = NULL, y = NULL, fill = "Score") +
  theme_minimal() + theme( panel.grid = element_blank())
  

print(p_v + p_u)  
}


```

