---
title: "Batch Effect Removed in u5, A26 and A27"
author: "Qiuyi Wu"
date: "`r Sys.Date()`"
output:
  html_document:
    toc: true
    toc_depth: 2
    toc_float: true
    code_folding: hide
---

```{r setup, include=FALSE}
library(Hmisc)
library(sRDA)
library(PMA)
library(corrplot)
library(dplyr)
library(readxl)
library(ggplot2)
library(ggtext)
library(pheatmap)
library(gridExtra)
library(factoextra)
library(kableExtra)
library(reshape2)
library(pROC)
library(randomForest)
library(glmnet)
library(DT)
library(reshape2)
library(sva)
library(writexl)
library(ggrepel)
library(purrr)
library(tidyverse)
library(plyr)
library(DT)
library(htmltools)
library(ggrepel)
library(sva)

knitr::opts_chunk$set(echo=TRUE, message=FALSE, cache=FALSE, warning=FALSE)
```



If we wan to compare A26, A27, U5, we need to remove the batch effect in these three datasets, so here we subtract the average of Q500 level 3 samples.


# U5

In this analysis, I proceed with 85% cutoff for the PCA analysis. Totally 622 metabolites, remove 22 metabolites that are 100% missing. After discussion, we decide to remove those missing > 70% for TG metabolites, and remove those missing > 85% for other metabolites. The number of metabolites reduced from 600 to **327**. 


Steps for PCA:

1. For TG metabolites, remove those missing > 70%; for other metabolites, remove those missing > 85%. 
2. Impute LOD/2
3. Dilution factor x0.222
5. Log transform



```{r load metabolomic u5}
# load FIA data
meta_met.u5fia <- read_excel("../../data/5583 FIA Q500 Data.xlsx", sheet = "5583 FIA Q500 Data", skip=1)
met_data.u5fia <- as.data.frame(meta_met.u5fia[c(6:80, 92:94), ])
# load UPLC data
meta_met.u5uplc <- read_excel("../../data/5583 UPLC Q500 Data.xlsx", sheet = "5583 UPLC Q500 Data", skip=1)
met_data.u5uplc <- as.data.frame(meta_met.u5uplc[c(6:80, 92:94), ])
# combine two data
met_data.u5 <- cbind(met_data.u5fia, met_data.u5uplc[, 12:219])
# remove the "status" columns
met_data.u5 <- met_data.u5[, !grepl("Status", colnames(met_data.u5))]
met_data.u5 <- cbind(met_data.u5$`Sample Identification`, met_data.u5[, -c(1:11)])
colnames(met_data.u5)[1] <- "Sample"
features.u5 <- colnames(met_data.u5)[-1]
# Replace all occurrences of "<LOD" with NA in the entire dataset
met_data.u5[met_data.u5 == "<LOD"] <- NA
# Convert all elements from character to numeric
met_data.u5[,-1] <- as.data.frame(lapply(met_data.u5[,-1], function(x) as.numeric(as.character(x))))
# remove 100% missing metabolites
met_data.u5 <- met_data.u5[, colMeans(is.na(met_data.u5)) < 1] # 22 metabolites with 100% missingness are removed, we have 600 metabolites left. 
```

```{r imput u5}
# Identify which columns are TG metabolites
tg_metabolites <- grepl("^TG", colnames(met_data.u5))
# Apply different missingness cutoffs
impute_cols <- colMeans(is.na(met_data.u5)) <= ifelse(tg_metabolites, 0.70, 0.85)
# Split data based on cutoff criteria
met_data.u5.impute <- met_data.u5[, impute_cols]  # 327 metabolites
met_data.u5.binary <- met_data.u5[, !impute_cols] # 273 metabolites

# impute metabolites 
met_data.u5.impute[,-1] <- lapply(met_data.u5.impute[,-1], function(col) {
  if (is.numeric(col)) {
    min_value <- min(col, na.rm = TRUE)
    col[is.na(col)] <- min_value / 2
  }
  return(col)
})
features <- colnames(met_data.u5.impute)[-1]

# binary metabolites 
met_data.u5.binary <- ifelse(is.na(met_data.u5.binary), 0, 1)

#  dilution factor 
met_data.u5.unscaling <- met_data.u5.impute
met_data.u5.unscaling[,-1] <- met_data.u5.impute[,-1]*0.222

# log2 transformation
met_data.u5.unscaling[,-1] <- log2(met_data.u5.unscaling[,-1] + 0.0001) 



# substract QC500 level 3 samples and remove them
qc3_u5 <- subset(met_data.u5.unscaling, Sample == "Quant 500_QC3")
qc3_u5means <- colMeans(qc3_u5[, features], na.rm = TRUE)
met_data.u5.unscaling[, features] <- sweep(
  met_data.u5.unscaling[, features],
  2,
  qc3_u5means,
  FUN = "-"
)
met_data.u5.unscaling <- subset(met_data.u5.unscaling, Sample != "Quant 500_QC3")


# Perform PCA on the data before Combat
met_unscaled_mat.u5 <- as.matrix(met_data.u5.unscaling[ , -1])  # remove metadata cols
pca_unscaled.u5 <- prcomp(met_unscaled_mat.u5, scale. = TRUE, center = FALSE)
pca_data_unscaled.u5 <- as.data.frame(pca_unscaled.u5$x[, 1:2]) 
colnames(pca_data_unscaled.u5) <- c("PC1", "PC2")
# Add sample info
pca_data_unscaled.u5$Sample <- met_data.u5.unscaling$Sample
# Mark outliers (beyond ±3 SD)
pca_data_unscaled.u5$z_PC1 <- scale(pca_data_unscaled.u5$PC1)
pca_data_unscaled.u5$z_PC2 <- scale(pca_data_unscaled.u5$PC2)
pca_data_unscaled.u5$outlier <- abs(pca_data_unscaled.u5$z_PC1) > 3 | abs(pca_data_unscaled.u5$z_PC2) > 3

ggplot(pca_data_unscaled.u5, aes(x = PC1, y = PC2)) +
  geom_point(size = 2) +
  geom_text_repel(
    data = subset(pca_data_unscaled.u5, outlier),
    aes(label = Sample),
    size = 3,
    max.overlaps = 100
  ) +
  labs(
    title = "PCA Plot of U5 Metabolome Data before ComBat",
    x = paste0("PC1 (", round(100 * summary(pca_unscaled.u5)$importance[2, 1], 1), "% variance)"),
    y = paste0("PC2 (", round(100 * summary(pca_unscaled.u5)$importance[2, 2], 1), "% variance)")
  ) +
  theme_minimal() +
  #scale_color_manual(values = c("cyan3", "red", "blue", "green", "purple", "orange", "pink")) +
  scale_shape_manual(values = c(16, 17))
```


# Bonnie A26

In this analysis, I proceed with 85% cutoff for the PCA analysis. Totally 630 metabolites, remove 13 metabolites that are 100% missing. After discussion, we decide to remove those missing > 70% for TG metabolites, and remove those missing > 85% for other metabolites. The number of metabolites reduced from 617 to **331**.  **Based on the result, we also decided to remove sample 244 (according to the 6/4/2025 group meeting)**. **Based on the downsteaming analysis, we also decided to remove sample 96 and 188 (according to the 8/20/2025 group meeting)**.

Steps for PCA:

1. only keep Q500(XL)_QC3 (remove Q500(XL)_QC1 and Q500(XL)_QC2 and SPQC samples)
2. Remove sample 244, 96, 188
3. For TG metabolites, remove those missing > 70%; for other metabolites, remove those missing > 85%. 
4. Impute LOD/2
5. Dilution factor x0.1
6. Log transform



```{r load metabolomic A26}
meta_met.a26 <- read_excel("../../data/10412-Q500 Data.xlsx", sheet = "1. 10412 Raw Data", skip=1)
# only keep Q500(XL)_QC3
met_data.a26 <- meta_met.a26[
  !meta_met.a26$`Sample identification` %in%
    c(
      grep("Q500\\(XL\\)_QC[12]$", meta_met.a26$`Sample identification`, value = TRUE),
      "101"
    ),
]
met_data.a26 <- cbind(met_data.a26$`Sample description`,  met_data.a26$`Submission name`, met_data.a26[, -c(1:11)])
colnames(met_data.a26)[1:2] <- c("Sample",  "Plate")
# Remove all the ratios/derived features before any analysis, 630 metabolites left
met_data.a26 <- met_data.a26[, 1:632] 
# use the same format ones as u5
features <- met_data.a26[1,-(1:2)]
met_data.a26 <- data.frame(met_data.a26[-c(1:14),])
met_data.a26$Sample[is.na(met_data.a26$Sample)] <- "Q500(XL)_QC3"
colnames(met_data.a26)[-(1:2)] <- features
# Convert all elements in met_data.a26 from character to numeric
met_data.a26[,-c(1:2)] <- as.data.frame(lapply(met_data.a26[,-c(1:2)], function(x) as.numeric(as.character(x))))


# Remove sample 244
met_data.a26 <- met_data.a26[!met_data.a26$Sample == "244",]
# Remove sample 96
met_data.a26 <- met_data.a26[!met_data.a26$Sample == "96",]
# Remove sample 188
met_data.a26 <- met_data.a26[!met_data.a26$Sample == "188",]

# Replace all occurrences of "<LOD" with NA in the entire dataset
met_data.a26[met_data.a26 == "<LOD"] <- NA
# Convert all elements in met_data.a26 from character to numeric
met_data.a26[,-c(1:2)] <- as.data.frame(lapply(met_data.a26[,-c(1:2)], function(x) as.numeric(as.character(x))))
# remove 100% missing metabolites
met_data.a26 <- met_data.a26[, colMeans(is.na(met_data.a26)) < 1] # 13 metabolites with 100% missingness are removed, we have 617 metabolites left. 
```



```{r imput a26}
# Identify which columns are TG metabolites
tg_metabolites <- grepl("^TG", colnames(met_data.a26))
# Apply different missingness cutoffs
impute_cols <- colMeans(is.na(met_data.a26)) <= ifelse(tg_metabolites, 0.70, 0.85)
# Split data based on cutoff criteria
met_data.a26.impute <- met_data.a26[, impute_cols]
met_data.a26.binary <- met_data.a26[, !impute_cols]


# impute metabolites 
met_data.a26.impute[,-c(1:2)] <- lapply(met_data.a26.impute[,-c(1:2)], function(col) {
  if (is.numeric(col)) {
    min_value <- min(col, na.rm = TRUE)
    col[is.na(col)] <- min_value / 2
  }
  return(col)
})
features <- colnames(met_data.a26.impute)[-c(1:2)]

# binary metabolites 
met_data.a26.binary <- ifelse(is.na(met_data.a26.binary), 0, 1)

# need to compare A22 and A26 scale -- dilution factor 
met_data.a26.unscaling <- met_data.a26.impute
met_data.a26.unscaling[,-c(1:2)] <- met_data.a26.impute[,-c(1:2)]*0.1

# log2 transformation
met_data.a26.unscaling[,-c(1:2)] <- log2(met_data.a26.unscaling[,-c(1:2)] + 0.0001)


# Compute per-plate QC means
qca26_rows <- grepl("Q500|QC", met_data.a26.unscaling$Sample)
qca26_means_per_plate <- met_data.a26.unscaling %>%
  filter(qca26_rows) %>%
  group_by(Plate) %>%
  dplyr::summarise(across(all_of(features), ~ mean(.x, na.rm = TRUE)))
# Subtract QC means
met_a26_adj <- met_data.a26.unscaling %>%
  left_join(qca26_means_per_plate, by = "Plate", suffix = c("", "_qcmean"))
for (f in features) {
  met_a26_adj[[f]] <- met_a26_adj[[f]] - met_a26_adj[[paste0(f, "_qcmean")]]
}
# Remove the columns with _qcmean
met_a26_adj <- met_a26_adj %>% select(-ends_with("_qcmean"))
# Remove QC samples 
met_data.a26.unscaling <- met_a26_adj %>% filter(!grepl("Q500|QC", Sample))



# Perform PCA on the data before Combat
met_unscaled_mat.a26 <- as.matrix(met_data.a26.unscaling[ , -c(1:2)])  # remove metadata cols
pca_unscaled.a26 <- prcomp(met_unscaled_mat.a26, scale. = TRUE, center = FALSE)
pca_data_unscaled.a26 <- as.data.frame(pca_unscaled.a26$x[, 1:2]) 
colnames(pca_data_unscaled.a26) <- c("PC1", "PC2")
# Add sample info
pca_data_unscaled.a26$Sample <- met_data.a26.unscaling$Sample
pca_data_unscaled.a26$color_indicator <- as.factor(met_data.a26.unscaling$Plate)
# Mark outliers (beyond ±3 SD)
pca_data_unscaled.a26$z_PC1 <- scale(pca_data_unscaled.a26$PC1)
pca_data_unscaled.a26$z_PC2 <- scale(pca_data_unscaled.a26$PC2)
pca_data_unscaled.a26$outlier <- abs(pca_data_unscaled.a26$z_PC1) > 3 | abs(pca_data_unscaled.a26$z_PC2) > 3

ggplot(pca_data_unscaled.a26, aes(x = PC1, y = PC2, color = color_indicator)) +
  geom_point(size = 2) +
  geom_text_repel(
    data = subset(pca_data_unscaled.a26, outlier),
    aes(label = Sample),
    size = 3,
    max.overlaps = 100
  ) +
  labs(
    title = "PCA Plot of A26 Metabolome Data before ComBat",
    x = paste0("PC1 (", round(100 * summary(pca_unscaled.a26)$importance[2, 1], 1), "% variance)"),
    y = paste0("PC2 (", round(100 * summary(pca_unscaled.a26)$importance[2, 2], 1), "% variance)")
  ) +
  theme_minimal() +
  scale_color_manual(values = c("cyan3", "red", "blue", "green", "purple", "orange")) +
  scale_shape_manual(values = c(16, 17))
```




# Frizzle A27

In this analysis, I proceed with 85% cutoff for the PCA analysis. Totally 630 metabolites, remove 10 metabolites that are 100% missing. After discussion, we decide to remove those missing > 70% for TG metabolites, and remove those missing > 85% for other metabolites. The number of metabolites reduced from 620 to **308**. 

Steps for PCA:

1. For TG metabolites, remove those missing > 70%; for other metabolites, remove those missing > 85%. 
2. Impute LOD/2
3. Dilution factor x0.1
4. Log transform



```{r load metabolomic a27}
meta_met.a27 <- read_excel("../../data/10788-Q500 Data.xlsx", sheet = "1. 10788 Raw Data", skip=1)
# only keep Q500(XL)_QC3
met_data.a27 <- meta_met.a27[
  !meta_met.a27$`Sample identification` %in%
    c(
      grep("Q500\\(XL\\)_QC[12]$", meta_met.a27$`Sample identification`, value = TRUE),
      "101"
    ),
]
met_data.a27 <- cbind(met_data.a27$`Sample description`,  met_data.a27$`Submission name`, met_data.a27[, -c(1:11)])
colnames(met_data.a27)[1:2] <- c("Sample", "Plate")
met_data.a27$Plate <- sub("^plate", "Plate", met_data.a27$Plate)

# Remove all the ratios/derived features before any analysis, 630 metabolites left
met_data.a27 <- met_data.a27[, 1:632] 
# use the same format ones as u5
features <- met_data.a27[1,-(1:2)]

met_data.a27 <- data.frame(met_data.a27[-c(1:121),])
met_data.a27$Sample[is.na(met_data.a27$Sample)] <- "Q500(XL)_QC3"
colnames(met_data.a27)[-(1:2)] <- features


# Replace all occurrences of "<LOD" with NA in the entire dataset
met_data.a27[met_data.a27 == "<LOD"] <- NA
# Convert all elements in met_data.a27 from character to numeric
met_data.a27[,-c(1:2)] <- as.data.frame(lapply(met_data.a27[,-c(1:2)], function(x) as.numeric(as.character(x))))
# remove 100% missing metabolites
met_data.a27 <- met_data.a27[, colMeans(is.na(met_data.a27)) < 1] # 10 metabolites with 100% missingness are removed, we have 620 metabolites left. 
```



```{r imput a27}
# Identify which columns are TG metabolites
tg_metabolites <- grepl("^TG", colnames(met_data.a27))
# Apply different missingness cutoffs
impute_cols <- colMeans(is.na(met_data.a27)) <= ifelse(tg_metabolites, 0.70, 0.85)
# Split data based on cutoff criteria
met_data.a27.impute <- met_data.a27[, impute_cols]  # 308 metabolites
met_data.a27.binary <- met_data.a27[, !impute_cols] # 312 metabolites


# impute metabolites 
met_data.a27.impute[,-c(1:2)] <- lapply(met_data.a27.impute[,-c(1:2)], function(col) {
  if (is.numeric(col)) {
    min_value <- min(col, na.rm = TRUE)
    col[is.na(col)] <- min_value / 2
  }
  return(col)
})
features <- colnames(met_data.a27.impute)[-c(1:2)]

# binary metabolites 
met_data.a27.binary <- ifelse(is.na(met_data.a27.binary), 0, 1)

# need to compare A22 and a27 scale -- dilution factor 
met_data.a27.unscaling <- met_data.a27.impute
met_data.a27.unscaling[,-c(1:2)] <- met_data.a27.impute[,-c(1:2)]*0.1

# log2 transformation
met_data.a27.unscaling[,-c(1:2)] <- log2(met_data.a27.unscaling[,-c(1:2)] + 0.0001) 


# Compute per-plate QC means
qca27_rows <- grepl("Q500|QC", met_data.a27.unscaling$Sample)
qca27_means_per_plate <- met_data.a27.unscaling %>%
  filter(qca27_rows) %>%
  group_by(Plate) %>%
  dplyr::summarise(across(all_of(features), ~ mean(.x, na.rm = TRUE)))
# Subtract QC means
met_a27_adj <- met_data.a27.unscaling %>%
  left_join(qca27_means_per_plate, by = "Plate", suffix = c("", "_qcmean"))
for (f in features) {
  met_a27_adj[[f]] <- met_a27_adj[[f]] - met_a27_adj[[paste0(f, "_qcmean")]]
}
# Remove the columns with _qcmean
met_a27_adj <- met_a27_adj %>% select(-ends_with("_qcmean"))
# Remove QC samples 
met_data.a27.unscaling <- met_a27_adj %>% filter(!grepl("Q500|QC", Sample))



# Perform PCA on the data before Combat
met_unscaled_mat.a27 <- as.matrix(met_data.a27.unscaling[ , -c(1:2)])  # remove metadata cols
pca_unscaled.a27 <- prcomp(met_unscaled_mat.a27, scale. = TRUE, center = FALSE)
pca_data_unscaled.a27 <- as.data.frame(pca_unscaled.a27$x[, 1:2]) 
colnames(pca_data_unscaled.a27) <- c("PC1", "PC2")
# Add sample info
pca_data_unscaled.a27$Sample <- met_data.a27.unscaling$Sample
pca_data_unscaled.a27$color_indicator <- as.factor(met_data.a27.unscaling$Plate)
# Mark outliers (beyond ±3 SD)
pca_data_unscaled.a27$z_PC1 <- scale(pca_data_unscaled.a27$PC1)
pca_data_unscaled.a27$z_PC2 <- scale(pca_data_unscaled.a27$PC2)
pca_data_unscaled.a27$outlier <- abs(pca_data_unscaled.a27$z_PC1) > 3 | abs(pca_data_unscaled.a27$z_PC2) > 3

ggplot(pca_data_unscaled.a27, aes(x = PC1, y = PC2, color = color_indicator)) +
  geom_point(size = 2) +
  geom_text_repel(
    data = subset(pca_data_unscaled.a27, outlier),
    aes(label = Sample),
    size = 3,
    max.overlaps = 100
  ) +
  labs(
    title = "PCA Plot of A27 Metabolome Data before ComBat",
    x = paste0("PC1 (", round(100 * summary(pca_unscaled.a27)$importance[2, 1], 1), "% variance)"),
    y = paste0("PC2 (", round(100 * summary(pca_unscaled.a27)$importance[2, 2], 1), "% variance)")
  ) +
  theme_minimal() +
  scale_color_manual(values = c("cyan3", "red", "blue", "green", "purple", "orange", "pink")) +
  scale_shape_manual(values = c(16, 17))
```

# Combined U5, A26 and A27
There are 252 metabolites for combined data of U5, A26 and A27. 


```{r Combined DATA}
# Merge u5, A26 and A27
common_cols <- Reduce(intersect, list(
  names(met_data.u5.unscaling),
  names(met_data.a26.unscaling),
  names(met_data.a27.unscaling)
)) #252 common metabolites

meta_met.u5_sub <- met_data.u5.unscaling[, common_cols]
meta_met.a26_sub <- met_data.a26.unscaling[, common_cols]
meta_met.a27_sub <- met_data.a27.unscaling[, common_cols]
met_data.comb <- rbind(meta_met.u5_sub, meta_met.a26_sub, meta_met.a27_sub)
met_data.comb$Subject <- c(rep("U5", nrow(meta_met.u5_sub)), rep("A26", nrow(meta_met.a26_sub)), rep("A27", nrow(meta_met.a27_sub)))

# Perform PCA on the combined data
met_combined_mat <- as.matrix(met_data.comb[ , -c(1, ncol(met_data.comb))])  
pca_combined <- prcomp(met_combined_mat, scale. = TRUE, center = TRUE)
# Extract first 2 PCs for plotting
pca_data.comb <- as.data.frame(pca_combined$x[, 1:2])
colnames(pca_data.comb) <- c("PC1", "PC2")
pca_data.comb$Sample <- met_data.comb$Sample
# Add Subject info to PCA data
pca_data.comb$Subject <- met_data.comb$Subject

pca_data.comb$z_PC1 <- scale(pca_data.comb$PC1)
pca_data.comb$z_PC2 <- scale(pca_data.comb$PC2)
pca_data.comb$outlier <- abs(pca_data.comb$z_PC1) > 3 | abs(pca_data.comb$z_PC2) > 3
pc1_var <- round(100 * summary(pca_combined)$importance[2, 1], 1)
pc2_var <- round(100 * summary(pca_combined)$importance[2, 2], 1)


ggplot(pca_data.comb, aes(x = PC1, y = PC2, color = Subject)) +
  geom_point(size = 2) +
  geom_text_repel(
    data = subset(pca_data.comb, outlier),
    aes(label = Sample),
    size = 3,
    max.overlaps = 100
  ) +
  labs(
    title = "PCA Plot of Combined Metabolome Data before ComBat",
    x = paste0("PC1 (", pc1_var, "% variance)"),
    y = paste0("PC2 (", pc2_var, "% variance)")
  ) +
  theme_minimal() +
  scale_color_manual(values = c("U5" = "purple", "A26" = "cyan3", "A27" = "red")) +
  scale_shape_manual(values = c(16, 17))
```


```{r PCA without outliers}
# Remove outliers
met_data.comb.clean <- met_data.comb[!met_data.comb$Sample %in% pca_data.comb$Sample[pca_data.comb$outlier], ]
met_combined_mat <- as.matrix(met_data.comb.clean[ , -c(1, ncol(met_data.comb.clean))])  
pca_combined <- prcomp(met_combined_mat, scale. = TRUE, center = TRUE)
# Extract first 2 PCs for plotting
pca_data.comb <- as.data.frame(pca_combined$x[, 1:2])
colnames(pca_data.comb) <- c("PC1", "PC2")
pca_data.comb$Sample <- met_data.comb.clean$Sample
# Add Subject info to PCA data
pca_data.comb$Subject <- met_data.comb.clean$Subject
pca_data.comb$z_PC1 <- scale(pca_data.comb$PC1)
pca_data.comb$z_PC2 <- scale(pca_data.comb$PC2)
pc1_var <- round(100 * summary(pca_combined)$importance[2, 1], 1)
pc2_var <- round(100 * summary(pca_combined)$importance[2, 2], 1)


# Plot without outliers
ggplot(pca_data.comb, aes(x = PC1, y = PC2, color = Subject)) +
  geom_point(size = 2) +
  labs(
    title = "PCA Plot of Combined Metabolome Data before ComBat  (Outliers Removed)",
    x = paste0("PC1 (", pc1_var, "% variance)"),
    y = paste0("PC2 (", pc2_var, "% variance)")
  ) +
  theme_minimal() +
  scale_color_manual(values = c("U5" = "purple", "A26" = "cyan3", "A27" = "red"))


loadings <- pca_combined$rotation[, 1]   # PC1 loadings
# Create dataframe of features + loadings
pc1_loadings_df <- data.frame(
  Feature = names(loadings),
  PC1_Loading = loadings,
  Abs_Loading = abs(loadings)
)
# Sort by magnitude
pc1_top_features <- pc1_loadings_df[order(-pc1_loadings_df$Abs_Loading), ]


top_n <- 20
ggplot(pc1_top_features[1:top_n, ],
       aes(x = reorder(Feature, PC1_Loading), y = PC1_Loading)) +
  geom_bar(stat = "identity") +
  coord_flip() +
  labs(
    title = paste("Top", top_n, "Features Driving PC1 Separation (U5/A26/A27)"),
    x = "Feature",
    y = "PC1 Loading"
  ) +
  theme_minimal()

datatable(pc1_top_features,
    options = list(pageLength = 10),
    caption = htmltools::tags$caption(
    style = 'caption-side: top; text-align: left; font-size: 120%; font-weight: bold;',
    "Features Driving PC1 Separation (U5/A26/A27)"))

```



# Combat

Now we consider all the U5/A26/A27 together with total 14 plates (A26 has 6 plates, A27 has 7 plates, and u5 itself is considered as 1 plate) and do Combat, to see if we can get rid of batch effect. 

```{r combat}
Plate <- c(rep("U5", nrow(met_data.u5.unscaling)), paste("A26", met_data.a26.unscaling$Plate), paste("A27", met_data.a27.unscaling$Plate))
met_data.comb.plt <- cbind(Plate, met_data.comb)



metabolite_data <- as.matrix(t(met_data.comb.plt[,3:254]))
met.batch <- as.factor(met_data.comb.plt$Plate)
# Apply ComBat to remove batch effect
combat_corrected <- ComBat(dat = metabolite_data, batch = met.batch, mod = NULL, par.prior = TRUE, prior.plots = FALSE)



pca_met_corrected <- prcomp(t(combat_corrected), scale. = TRUE)
pca_data <- data.frame(pca_met_corrected$x)  # PCA scores
pca_data$Plate <- met_data.comb.plt$Plate
pca_data$Subject <- met_data.comb.plt$Subject

ggplot(pca_data, aes(PC1, PC2, color = Plate)) +
  geom_point(size = 2) +
  theme_minimal() +
  labs(title = "PCA After ComBat Plate Correction")

ggplot(pca_data, aes(PC1, PC2, color = Subject)) +
  geom_point(size = 2) +
  theme_minimal() +
  labs(title = "PCA After ComBat Plate Correction")




loadings <- pca_met_corrected$rotation[, 1]   # PC1 loadings
# Create dataframe of features + loadings
pc1_loadings_df <- data.frame(
  Feature = names(loadings),
  PC1_Loading = loadings,
  Abs_Loading = abs(loadings)
)
# Sort by magnitude
pc1_top_features <- pc1_loadings_df[order(-pc1_loadings_df$Abs_Loading), ]


top_n <- 20
ggplot(pc1_top_features[1:top_n, ],
       aes(x = reorder(Feature, PC1_Loading), y = PC1_Loading)) +
  geom_bar(stat = "identity") +
  coord_flip() +
  labs(
    title = paste("Top", top_n, "Features Driving PC1 Separation (U5/A26/A27)"),
    x = "Feature",
    y = "PC1 Loading"
  ) +
  theme_minimal()

datatable(pc1_top_features,
    options = list(pageLength = 10),
    caption = htmltools::tags$caption(
    style = 'caption-side: top; text-align: left; font-size: 120%; font-weight: bold;',
    "Features Driving PC1 Separation (U5/A26/A27)"))

```
# Two Sample Testing

Here we conducted two-sample testing (t-test) between u5 and the A26/A27 on the visual cortex area for each of the metabolite. We computed adjusted p-values using the Benjamini–Hochberg (BH) procedure, which controls the false discovery rate (FDR).


```{r two sample test}
a26.v1 <- data.frame(read_excel("../../data/A26_Bonnie_Flatmap_Grid_20251204.xlsx", sheet = "Master_Decoder"))
a26.v1 <- a26.v1[,c("Analysis.ID", "ContainsV1")]
a26.v1 <- a26.v1$Analysis.ID[a26.v1$ContainsV1 %in% "Yes"]


a27.v1 <- data.frame(read_excel("../../data/A27_Frizzle_Flatmap_Grid_20251204.xlsx", sheet = "Decoder"))
a27.v1 <- a27.v1[,c("AnalysisID", "ContainsV1")]
a27.v1 <- a27.v1$AnalysisID[a27.v1$ContainsV1 %in% "Yes"]

u5.v1 <- data.frame(read_excel("../../data/U6_NotExsanguinated_Samples.xlsx", sheet = "U6"))
u5.v1 <- u5.v1[,c("Duke.ID", "ContainsV1")]
u5.v1 <- u5.v1$Duke.ID[u5.v1$ContainsV1 %in% "Yes"]

met.comb.correct <- data.frame(
  Subject = met_data.comb.plt$Subject,
  Sample  = met_data.comb.plt$Sample,
  t(combat_corrected)
)

# visual cortex area samples
met.comb.v1subset <- met.comb.correct[(met.comb.correct$Subject == "U5" & met.comb.correct$Sample %in% u5.v1) | (met.comb.correct$Subject == "A26" & met.comb.correct$Sample %in% a26.v1) | (met.comb.correct$Subject == "A27" & met.comb.correct$Sample %in% a27.v1),]

# Identify metabolite columns 
met_cols <- setdiff(colnames(met.comb.v1subset), c("Subject", "Sample"))

# Create group variable: u5 vs A26/A27
group <- ifelse(met.comb.v1subset$Subject == "U5", "U5", "A26A27")

# Initialize results
ttest_results <- data.frame(
  Metabolite = met_cols,
  t_stat = NA,
  p_value = NA,
  mean_u5 = NA,
  mean_A26A27 = NA
)

for (i in seq_along(met_cols)) {
  metabolite <- met_cols[i]
  x <- met.comb.v1subset[group == "U5", metabolite]
  y <- met.comb.v1subset[group == "A26A27", metabolite]
  
  # t-test 
  tt <- t.test(x, y)
  
  # store
  ttest_results$t_stat[i]     <- round(tt$statistic, 4)
  ttest_results$p_value[i]    <- tt$p.value
  ttest_results$mean_u5[i]    <- round(mean(x, na.rm = TRUE), 4)
  ttest_results$mean_A26A27[i] <- round(mean(y, na.rm = TRUE), 4)
}

# Add FDR correction
ttest_results$adj.p_value <- p.adjust(ttest_results$p_value, method = "BH")

# Sort by p-value
ttest_results <- ttest_results[order(ttest_results$adj.p_value), ]

datatable(ttest_results,
    options = list(pageLength = 10),
    caption = htmltools::tags$caption(
    style = 'caption-side: top; text-align: left; font-size: 120%; font-weight: bold;',
    "Two-sample t-tests for each metabolite (U5 vs A26/A27)"))

```

Here we conducted two-sample testing (t-test) for U5 vs A26 and u5 vs A27 on the visual cortex area for each of the metabolite. We computed adjusted p-values using the Benjamini–Hochberg (BH) procedure, which controls the false discovery rate (FDR).

We also plot the -log10 p-values of the u5 vs A26 comparison as x-axis, and the -log10 p-values of the u5 vs A27 comparison as y-axis to see if the comparisons are consistent. 


```{r two sample test seperately}
# compare u5 and A26
met.comb.v1subset.u5a26 <- met.comb.v1subset[met.comb.v1subset$Subject != "A27",]
# Create group
group <- met.comb.v1subset.u5a26$Subject 

# Initialize results
ttest_results.u5a26 <- data.frame(
  Metabolite = met_cols,
  t_stat = NA,
  p_value = NA,
  mean_u5 = NA,
  mean_A26 = NA
)

for (i in seq_along(met_cols)) {
  metabolite <- met_cols[i]
  x <- met.comb.v1subset.u5a26[group == "U5", metabolite]
  y <- met.comb.v1subset.u5a26[group == "A26", metabolite]
  
  # t-test 
  tt <- t.test(x, y)
  
  # store
  ttest_results.u5a26$t_stat[i]     <- round(tt$statistic, 4)
  ttest_results.u5a26$p_value[i]    <- tt$p.value
  ttest_results.u5a26$mean_u5[i]    <- round(mean(x, na.rm = TRUE), 4)
  ttest_results.u5a26$mean_A26[i] <- round(mean(y, na.rm = TRUE), 4)
}

# Add FDR correction
ttest_results.u5a26$adj.p_value <- p.adjust(ttest_results.u5a26$p_value, method = "BH")

# Sort by p-value
ttest_results.u5a26 <- ttest_results.u5a26[order(ttest_results.u5a26$adj.p_value), ]

datatable(ttest_results.u5a26,
    options = list(pageLength = 10),
    caption = htmltools::tags$caption(
    style = 'caption-side: top; text-align: left; font-size: 120%; font-weight: bold;',
    "Two-sample t-tests for each metabolite (U5 vs A26)"))



# compare u5 and A27
met.comb.v1subset.u5a27 <- met.comb.v1subset[met.comb.v1subset$Subject != "A26",]
# Create group
group <- met.comb.v1subset.u5a27$Subject 

# Initialize results
ttest_results.u5a27 <- data.frame(
  Metabolite = met_cols,
  t_stat = NA,
  p_value = NA,
  mean_u5 = NA,
  mean_A27 = NA
)

for (i in seq_along(met_cols)) {
  metabolite <- met_cols[i]
  x <- met.comb.v1subset.u5a27[group == "U5", metabolite]
  y <- met.comb.v1subset.u5a27[group == "A27", metabolite]
  
  # t-test 
  tt <- t.test(x, y)
  
  # store
  ttest_results.u5a27$t_stat[i]     <- round(tt$statistic, 4)
  ttest_results.u5a27$p_value[i]    <- tt$p.value
  ttest_results.u5a27$mean_u5[i]    <- round(mean(x, na.rm = TRUE), 4)
  ttest_results.u5a27$mean_A27[i] <- round(mean(y, na.rm = TRUE), 4)
}

# Add FDR correction
ttest_results.u5a27$adj.p_value <- p.adjust(ttest_results.u5a27$p_value, method = "BH")

# Sort by p-value
ttest_results.u5a27 <- ttest_results.u5a27[order(ttest_results.u5a27$adj.p_value), ]

datatable(ttest_results.u5a27,
          options = list(pageLength = 10),
          caption = htmltools::tags$caption(
            style = 'caption-side: top; text-align: left; font-size: 120%; font-weight: bold;',
            "Two-sample t-tests for each metabolite (U5 vs A27)"))



pval_compare <- merge(
  ttest_results.u5a26[, c("Metabolite", "p_value")],
  ttest_results.u5a27[, c("Metabolite", "p_value")],
  by = "Metabolite",
  suffixes = c("_u5a26", "_u5a27")
)

pval_compare$log10_u5a26 <- -log10(pval_compare$p_value_u5a26)
pval_compare$log10_u5a27 <- -log10(pval_compare$p_value_u5a27)

ggplot(pval_compare,
       aes(x = log10_u5a26, y = log10_u5a27)) +
  geom_point(alpha = 0.6, size = 1.5) +
  geom_abline(slope = 1, intercept = 0, linetype = "dashed", color = "gray50") +
  theme_minimal() +
  labs(
    title = "Consistency of Differential Signals",
    subtitle = "T-test (U5 vs A26) vs (U5 vs A27)",
    x = expression(-log[10](p[u5~vs~A26])),
    y = expression(-log[10](p[u5~vs~A27]))
  )
```

Trying Wilcoxon rank sum test for two-sample testing. 

```{r w-test}
# compare u5 and A26
met.comb.v1subset.u5a26 <- met.comb.v1subset[met.comb.v1subset$Subject != "A27",]
# Create group
group <- met.comb.v1subset.u5a26$Subject 

# Initialize results
wtest_results.u5a26 <- data.frame(
  Metabolite = met_cols,
  w_stat = NA,
  p_value = NA,
  mean_u5 = NA,
  mean_A26 = NA
)

for (i in seq_along(met_cols)) {
  metabolite <- met_cols[i]
  x <- met.comb.v1subset.u5a26[group == "U5", metabolite]
  y <- met.comb.v1subset.u5a26[group == "A26", metabolite]
  
  # w-test 
  wt <- wilcox.test(x, y, exact = FALSE)
  
  # store
  wtest_results.u5a26$t_stat[i]     <- round(wt$statistic, 4)
  wtest_results.u5a26$p_value[i]    <- wt$p.value
  wtest_results.u5a26$mean_u5[i]    <- round(mean(x, na.rm = TRUE), 4)
  wtest_results.u5a26$mean_A26[i] <- round(mean(y, na.rm = TRUE), 4)
}

# Add FDR correction
wtest_results.u5a26$adj.p_value <- p.adjust(wtest_results.u5a26$p_value, method = "BH")

# Sort by p-value
wtest_results.u5a26 <- wtest_results.u5a26[order(wtest_results.u5a26$adj.p_value), ]

datatable(wtest_results.u5a26,
          options = list(pageLength = 10),
          caption = htmltools::tags$caption(
            style = 'caption-side: top; text-align: left; font-size: 120%; font-weight: bold;',
            "Two-sample w-tests for each metabolite (U5 vs A26)"))



# compare u5 and A27
met.comb.v1subset.u5a27 <- met.comb.v1subset[met.comb.v1subset$Subject != "A26",]
# Create group
group <- met.comb.v1subset.u5a27$Subject 

# Initialize results
wtest_results.u5a27 <- data.frame(
  Metabolite = met_cols,
  w_stat = NA,
  p_value = NA,
  mean_u5 = NA,
  mean_A27 = NA
)

for (i in seq_along(met_cols)) {
  metabolite <- met_cols[i]
  x <- met.comb.v1subset.u5a27[group == "U5", metabolite]
  y <- met.comb.v1subset.u5a27[group == "A27", metabolite]
  
  # w-test 
  wt <- wilcox.test(x, y, exact = FALSE)
  
  # store
  wtest_results.u5a27$w_stat[i]     <- round(wt$statistic, 4)
  wtest_results.u5a27$p_value[i]    <- wt$p.value
  wtest_results.u5a27$mean_u5[i]    <- round(mean(x, na.rm = TRUE), 4)
  wtest_results.u5a27$mean_A27[i] <- round(mean(y, na.rm = TRUE), 4)
}

# Add FDR correction
wtest_results.u5a27$adj.p_value <- p.adjust(wtest_results.u5a27$p_value, method = "BH")

# Sort by p-value
wtest_results.u5a27 <- wtest_results.u5a27[order(wtest_results.u5a27$adj.p_value), ]

datatable(wtest_results.u5a27,
          options = list(pageLength = 10),
          caption = htmltools::tags$caption(
            style = 'caption-side: top; text-align: left; font-size: 120%; font-weight: bold;',
            "Two-sample w-tests for each metabolite (U5 vs A27)"))



pval_compare <- merge(
  wtest_results.u5a26[, c("Metabolite", "p_value")],
  wtest_results.u5a27[, c("Metabolite", "p_value")],
  by = "Metabolite",
  suffixes = c("_u5a26", "_u5a27")
)

pval_compare$log10_u5a26 <- -log10(pval_compare$p_value_u5a26)
pval_compare$log10_u5a27 <- -log10(pval_compare$p_value_u5a27)

ggplot(pval_compare,
       aes(x = log10_u5a26, y = log10_u5a27)) +
  geom_point(alpha = 0.6, size = 1.5) +
  geom_abline(slope = 1, intercept = 0, linetype = "dashed", color = "gray50") +
  theme_minimal() +
  labs(
    title = "Consistency of Differential Signals",
    subtitle = "W-test (U5 vs A26) vs (U5 vs A27)",
    x = expression(-log[10](p[u5~vs~A26])),
    y = expression(-log[10](p[u5~vs~A27]))
  )
```


